Verification of Stencil Computations through Spatial Specifications

### 2016-07-03 paper.tex
fixed the definition of effective dimensions
merge-conflict

In this work, we introduce a simple specification language for the
spatial properties of stencil computations. The
specifications abstract over the fine grained detail of stencil access
patterns.  In practise, most stencil computations have a regular shape
that can be described simply and abstractly, with a small set of
coarse-grained descriptions. In the case of our first Laplace example,
our inference procedure provides the specification:
%
\begin{SpecVerbatim}
!=  stencil centered(depth=1, dim=1) :: a
\end{SpecVerbatim}
%
This explains that \texttt{a} is read from with a
symmetrical stencil pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Figure~\ref{ref:navier-stokes-fragment}(c) describes these
pictorially.

@@

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+
          (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{ExmVerbatim}
(a). Excerpt from Fortran code for a Navier-Stoke fluid simulator,
showing highly-detailed stencil computations. \\

@@

\begin{SpecVerbatim}[xleftmargin=0.3cm]
!=   stencil centered(depth=1, dim=1)
           + centered(depth=1, dim=2) :: u

!=   stencil forward(depth=1, dim=1)
           * backward(depth=1, dim=2) :: v
\end{SpecVerbatim}
(b). Inferred stencil specification from CamFort

\begin{center}
\begin{tikzpicture}
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
(c). Pictorial representation of the two stencil specifications.
The horizontal dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
\caption{Fragment of Navier-Stokes fluid simulator and its specification}
\label{ref:navier-stokes-fragment}
\end{figure}

@@

Regions have as terminals the \emph{region constants}, denoted by
\term{forward}, \term{backward}, or \term{centered}. Each
region constant is given a depth parameter $n$ (natural number greater
than 0) and a dimension identifier $d$ (also a natural number greater than
0). Region constants specify that an array is read from by a collection
of indices which in the $d^{th}$ dimension are
neighbourhood indices ranging from $i + 0$ up to $i + n$ inclusively, \eg{}, the
following is valid stencil, reading from \term{a} and writing to \term{b}:
%%
\begin{ExmVerbatim}
!= stencil forward(depth=2, dim=1) :: a
b(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
\end{ExmVerbatim}
%%
were \texttt{i} is an induction variable.  The stencil specification
is associated to the array variable \texttt{a}. Note that the second
dimension is fixed at a constant here.

@@

The sum of two regions $r \texttt{+} r'$ specifies that an array is
read using the neighbourhood indices described by both $r$ and
$r'$. It can be thought of as a conjunction of specifications. For
example, the following gives the specification for the five-point
stencil which is the sum of two \texttt{centered} regions in each
dimension:
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1)
         + centered(depth=1, dim=2) :: a

>>>>

In this work, we introduce a simple specification language for the
spatial properties of stencil computations. The
specifications abstract over the fine grained detail of stencil access
patterns. In practice, most stencil computations have a regular shape
that can be described simply and abstractly, with a small set of
coarse-grained descriptions. In the case of our first Laplace example,
our inference procedure provides the specification:
%
\begin{SpecVerbatim}
!=  stencil centered(depth=1, dim=1) :: a
\end{SpecVerbatim}
%
This explains that \texttt{a} is read from with a
symmetrical stencil pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Figure~\ref{ref:navier-stokes-fragment}(c) describes these
pictorially.

@@

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+
          (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely
\end{ExmVerbatim}
(a) Excerpt from Fortran code for a Navier-Stoke fluid simulator,
showing highly-detailed stencil computations. \\

@@

\begin{SpecVerbatim}[xleftmargin=0cm]
!= stencil centered(depth=1, dim=1)*reflexive(dim=2)
         + centered(depth=1, dim=2)*reflexive(dim=1) :: u

!= stencil forward(depth=1, dim=1)
         * backward(depth=1, dim=2) :: v
\end{SpecVerbatim}
(b) Inferred stencil specification from CamFort

\begin{center}
\begin{tikzpicture}
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
(c) Pictorial representation of the two stencil specifications.
The horizontal dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
\caption{Fragment of Navier-Stokes fluid simulator and its specification}
\label{ref:navier-stokes-fragment}
\end{figure}

@@

Regions have as terminals the \emph{region constants}, denoted by
\term{reflexive}, \term{forward}, \term{backward}, or \term{centered}. Each
region constant except \term{reflexive} is given a depth parameter $n$ (natural
number greater than 0). They all receive a dimension identifier $d$ (also a
natural number greater than 0). Region constants specify that an array is read
from by a collection of indices which in the $d^{th}$ dimension are
neighbourhood indices ranging from $i + 0$ up to $i \pm n$ inclusively,
\eg{}, the following is valid stencil, reading from \term{a} and writing to
\term{b}:
%%
\begin{ExmVerbatim}
!= stencil forward(depth=2, dim=1) :: a
b(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
\end{ExmVerbatim}
%%
were \texttt{i} is an induction variable.  The stencil specification
is associated to the array variable \texttt{a}. Note that the second
dimension is fixed at a constant here.

@@

The sum of two regions $r \texttt{+} r'$ specifies that an array is
read using the neighbourhood indices described by either $r$ or
$r'$. It can be thought of as a disjunction of specifications. For
example, the following gives the specification for the five-point
stencil which is the sum of two \texttt{centered} regions in each
dimension:
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1)*reflexive(dim=2)
         + centered(depth=1, dim=2)*reflexive(dim=1) :: a
